---
title: 'Course Project: Demonstrations of the central limit theorem and using the
  t distribution in hypothesis testing'
author: "Vineet W. Singh"
date: "19 February 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Introduction

Part 1 of this project, briefly demonstrates the central limit theorem and the
law of large numbers at work while using about 1000 samples.  

Part 2 demonstrates the use of the t distribution in finding confidence intervals 
for a two sets of data of about 30 observations each.  

### Part 1

This part demonstrates the law of large numbers and the central limit theorem. 
The law of large numbers states that given enough observations, any statistics 
generated from those observations will estimate the pouplation statistics with 
some accuracy.  

For example, given a data set with a large number of observations/sampling units, 
that has been collected from a population that is exponentially distibuted and 
which has a certain rate parameter(lambda), then the theoretical population mean 
is supposed to be equal to 1/lambda and the theoretical standard deviation of 
the mean is also supposed to be equal to 1/lambda.  

In this chunk of R code, we randomly generate 1000 observations from an 
exponential distribution with a rate parameter of 0.2.  

```{r chunk1}
mexp<-rexp(1000,0.2)
```

The mean of this particular sample of 1000 observations is:
```{r chunk2}
print(mean(mexp))
```
The theoretical mean of the original population that follows the exponential 
distribution is 1/lambda= 1/0.2 = 5.  

As demonstrated, the mean from the randomly generated set is quite close to 
that of the population mean and can be taken to be a fair/close estimate of the
population mean.  

A histogram with the original distribution values and the theoretical mean, 
sample mean is displayed below:  
```{r chunk 3}
hist(mexp, main=paste0("Histogram of 1000 randomly genererated points\n",
                        "from an exponential distribution "),
     xlab="Sample Values")
abline(v=mean(mexp),col=2,lwd=4)
abline(v=5,col=4,lwd=4)
legend("topright", c("Sample Mean","Theoretical Mean"), col=c(2,4), lwd=4)
```

The central limit theorem states that the means of a large number of samples
will be normally distributed around the theoretical population mean.  

To demonstrate this we first generate 1000 samples of 1000 
observations, randomly taken from the exponential distribution with a 
population rate parameter(lambda) of 0.2.  

As demonstrated above the population mean in supposed to be 5 and the 
standard deviation of the population mean is also supposed to be 5.    

```{r chunk4}
#mns is a list containing 1000 mean values of 1000 observations randomly
#generated from a exponential distribution
mns=NULL
#sds is a list containing 1000 standard deviation of mean of 1000 observations 
#randomly generated from the same data as above.
sds=NULL
for (i in 1 : 1000){
x<-rexp(1000,0.2)
mns=c(mns,mean(x))
sds=c(sds,sd(x))
}
```

We can now plot a simple histogram to observe how the actual means from the 
1000 samples are distributed. The theoretical population mean is 
demonstrated as a vertical red line:

```{r chunk5}
hist(mns, main=paste0("Histogram of means of 1000 samples\n",
                       "each sample has 1000 observations\n",
                    "from a population following an exponential distribution"),
     xlab="Mean") 
abline(v=5,col=2,lwd=4)
legend("topright", "Theoretical Mean", col=2, lwd=4)
```

As can be seen from the histogram above, the generated distribution is fairly 
symmetrical around the theoretical population mean.  

Further, there are two ways to demonstrate that the sample means are normally 
distributed.  

One is to convert the frequencies of the means into  probabilities and 
plot this data as a histogram and superimpose onto this histogram, 
a density line representing probability density:
```{r chunk6}
hist(mns, main=paste0("Histogram of means of 1000 samples\n",
                       "each sample has 1000 observations\n",
                    "from a population following an exponential distribution"),
     xlab="Mean",
     probability = TRUE) 
lines(density(mns,adjust=2),col=4)
abline(v=5,col=2,lwd=4)
legend("topright", c("Theoretical Mean","Probability Density"),col=c(2,4),lwd=4)
```

Another way is to use the qqnorm function to generate a log normal plot of the 
data. If the plot points fall on a straight line the data follows a normal 
distribution.  

```{r chunk7}
qqnorm(mns, main=paste0("Log Normal Plot of means of 1000 samples\n",
                       "each sample has 1000 observations\n",
                    "from a population following an exponential distribution"))
qqline(mns,col=2)
legend("bottomright", "Log Normal Line",col=2,lwd=4)
```

Similarly we can prove that the standard deviations also are normally 
distributed by  

1) Showing that the standard deviations are symmetrerical around the 
population SD.  

```{r chunk8}
hist(sds, main=paste0("Histogram of Standard Deviation of means of \n",
                       "1000 samples, each sample has 1000 observations\n",
                    "from a population following an exponential distribution"),
     xlab="SD") 
abline(v=5,col=2,lwd=4)
legend("topright", "Theoretical SD", col=2, lwd=4)
```

and  

2) By plotting the log normal plot of the SD values.  

```{r chunk9}
qqnorm(sds, main=paste0("Log Normal Plot of SD of means of 1000 samples\n",
                       "each sample has 1000 observations\n",
                    "from a population following an exponential distribution"))
qqline(sds,col=2)
legend("bottomright", "Log Normal Line",col=2,lwd=4)
```


### Conclusion  

We have shown as to how the central limit theorem is satisfied for both the 
mean and the standard deviation of simulated 1000 samples. The original 
population is exponentially distributed but the means and SD's of all samples
generated from this exponential distribution follow a normal distribution 
centered at the population mean and SD respectively. 

## Part 2

Part 2 of this project demostrates the use of use of the t distribution to 
generate confidence intervals and t values to do hypothesis testing based on 
the ToothGrowth data set that is a part of the datasets library in R. 

We hope to answer the following questions after analysing the data:  

1) Do the teeth grow the same, for all dosage of supplements or do different 
doses of supplements result in different growth.   

2) Do both the supplements result in the same rate of growth of teeth or is 
the growth rate different for each supplement.

The tooth growth data set contains data regarding Vitamin C given as a
supplement and the resulting growth in teeth of 60 guinea pigs. 
More information about the dataset can be found at: 
https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html

### Analysis

To begin the analysis the dplyr and datasets packages are loaded using the 
library command. This automatically loads the tooth growth data into R. 

```{r chunk10}
library(datasets)
library(dplyr)
```

The structure of the ToothGrowth dataset can be found using the str function.  

```{r chunk11}
str(ToothGrowth)
```

We need to partition this set according to the supplements concerned, either
"VC" for Ascorbic Acid or "OJ" for orange juice. The resultant datasets are 
stored in different frames.

We also need to add a new variable "rate" which will measure the rate at which 
the teeth have grown according to the amount of supplement given. In effect, we 
are normalising the length with respect to the dose. 

```{r chunk12}
tg1 <- ToothGrowth %>% filter(supp=="VC")
tg1<-tg1 %>% mutate(rate=len/dose)
tg2 <- ToothGrowth %>% filter(supp=="OJ")
tg2<-tg2 %>% mutate(rate=len/dose)
```

We can also generate some statistics for each supplement group. 

```{r chunk13}
#calculate summaries for each sub group of the two groups 
tg1 %>% group_by(dose) %>% summarise(mean=mean(rate),sd=sd(rate),n=n())
tg2 %>% group_by(dose) %>% summarise(mean=mean(rate),sd=sd(rate),n=n())
```
The growth rates according to the supplement type and dosage groups are:

#### Ascorbic Acid : Summary 
<table style="border: 1px solid black;">
<tr>
  <th style="border: 1px solid black;"> Dose(mg/day)  </td>
  <th style="border: 1px solid black;"> Mean</td>
  <th style="border: 1px solid black;"> SD </td>
</tr>
<tr>
  <td style="border: 1px solid black;"> 0.5</td>
  <td style="border: 1px solid black;">16</td>
  <td style="border: 1px solid black;">5.49</td>
</tr>
<tr>
  <td style="border: 1px solid black;">1</td>
  <td style="border: 1px solid black;">16.8</td>
  <td style="border: 1px solid black;">2.52</td>
</tr>
<tr>
  <td style="border: 1px solid black;">2</td>
  <td style="border: 1px solid black;">13.1</td>
  <td style="border: 1px solid black;">2.40</td>
</tr>
</table>
  
  
#### Orange Juice : Summary 
<table style="border: 1px solid black;">
<tr>
  <th style="border: 1px solid black;"> Dose(mg/day)  </td>
  <th style="border: 1px solid black;"> Mean</td>
  <th style="border: 1px solid black;"> SD </td>
</tr>
<tr>
  <td style="border: 1px solid black;"> 0.5</td>
  <td style="border: 1px solid black;">26.5</td>
  <td style="border: 1px solid black;">8.92</td>
</tr>
<tr>
  <td style="border: 1px solid black;">1</td>
  <td style="border: 1px solid black;">22.7</td>
  <td style="border: 1px solid black;">3.91</td>
</tr>
<tr>
  <td style="border: 1px solid black;">2</td>
  <td style="border: 1px solid black;">13</td>
  <td style="border: 1px solid black;">1.33</td>
</tr>
</table>

We can draw a plot of the growth rates according to the supplement types. 

```{r chunk14}
plot(x=row.names(tg1),y=tg1$rate,type="p",pch=1,ylab="Rate",
     main="Tooth Growth Rate According to Supplement",
     xlab="Observation No.", ylim=c(0,50),col=1)
lines(x=row.names(tg1),y=tg1$rate,type="l",lty=1,lwd=2,col=1)
points(x=row.names(tg2),y=tg2$rate,pch=1,col=2)
lines(x=row.names(tg2),y=tg2$rate,type="l",lty=1,lwd=2,col=2)
legend("topright", c("Ascorbic Acid","Orange Juice"),col=c(1,2),lwd=4)
```

From this preliminary analysis, we can observe that the growth rates for both 
supplements are different. 
Within each supplement group the growth rates vary according to dosages of 
supplement given, with the maximum variabilty in the 0.5 mg group for 
both supplements. 

But to get detailed answers we will need to formulate some hypothesis and 
carry out some tests. 

We will need to make some assumptions about the data before we carry out the 
tests. 

We can assume that the growth rate in each animal in each subgroup is 
independent since each animal is different. 
Further more we can assume that the variance 
within each of the 6 subgroups (supplement & dosage) is the same. 

However, it is probably not very wise to assume that the variance accross 
different subgroups is the same as, clearly the dosage levels are different 
across each subgroup in one supplement group. 

The first null hypothesis we propose is:  

#### Null Hypothesis No. 1
Mean growth rates for the 0.5(mg/day) group and the 1(mg/day) group of the 
ascorbic acid set are the same.

To assist us in our analysis we divide the data up further into three groups for 
each supplement type. 

```{r chunk15}
tg11<- tg1 %>% filter(dose==0.5)
tg12<- tg1 %>% filter(dose==1)
tg13<- tg1 %>% filter(dose==2)

tg21<- tg2 %>% filter(dose==0.5)
tg22<- tg2 %>% filter(dose==1)
tg23<- tg2 %>% filter(dose==2)
```

Since we only have 10 measurements each subgroup, we need to make 
confidence intervals and do tests using the t distribution. 
This is done using the following command: 

```{r chunk16}
t.test(tg11$rate,tg12$rate,paired=F,var.equal = F)
```

The t-value is -0.42396, p-value is above .50 and the confidence interval for 
the difference of means  
contains 0. 

<u> <b>We cannot, therefore, reject null hypothesis no. 1.</b> </u>  

We can assume that the mean growth rates for both 0.5 (mg/day) and 1 (mg/day) 
dose groups of the ascorbic acid supplement set are the same. 

The next null hypothesis we propose is: 

#### Null Hypothesis No. 2
Mean growth rates for the 1(mg/day) group and the 2(mg/day) group of the 
ascorbic acid set are the same.

We test using the t.test function as outlined above: 

```{r chunk17}
t.test(tg12$rate,tg13$rate,paired=F,var.equal = F)
```

Here we come across a very high t-value of 3.3662, a low p value and a 
confidence interval where the difference is well above 1.  
<b> <u> We can, therefore, 
reject the null hypothesis no. 2 </u> </b> and conclude that the means for 
the two subgroups 
are different.  

We propose the following hypotheses:  

#### Null hypothesis No. 3: 
Mean growth rates for the 0.5(mg/day) group and the 1(mg/day) group of the 
orange juice set are the same.

#### Null hypothesis No. 4
Mean growth rates for the 1(mg/day) group and the 2(mg/day) group of the 
orange juice set are the same.

```{r chunk18}
t.test(tg21$rate,tg22$rate,paired=F,var.equal = F)
t.test(tg22$rate,tg23$rate,paired=F,var.equal = F)
```

With the test number three, we find that the t-value is above 1 but not 
very high, the confidence interval of the difference contains 0 so we cannot 
rule out that the means might not be the same. Results of test number 4 tell a 
different story, however, the t-value is very large, the p-value is negligible 
and the confidence interval of the difference does not contain 0.  

<b> <u> We cannot, therefore, reject null hypothesis no. 3, but using the 
argument in the sentence above, we can reject null hypothesis no. 4. </u> </b>

Based on the results of the 4 hypothesis tests above, we should merge the two 
groups of 0.5(mg/day) and 1(mg/day) for each of the two supplement types. 
and then do a comparision of the two supplement types based on two groups for 
each of the supplement type 0.5-1(mg/day) and 2(mg/day).

```{r chunk19}
#CI for mean growth rate for tg11 and tg12 includes 0 so we cannot rule out
#that the mean growth rate for both tg21 and tg22 can be the same
#it might make sense to group tg11 and tg12 together and assume a common mean 
#growth rate. 

#merge tg21 and tg22 into one frame
tg11<-rbind(tg11,tg12)

#reassign tg23 to tg22
tg12<-tg13

#CI for mean growth rate for tg21 and tg22 includes 0 so we cannot rule out
#that the mean growth rate for both tg21 and tg22 can be the same
#it might make sense to group tg21 and tg22 together and assume a common growth 
#rate. 

#merge tg21 and tg22 into one frame
tg21<-rbind(tg21,tg22)

#reassign tg23 to tg22
tg22<-tg23
```

The new data sets are ready for the final phase of the analysis. 
Hypothesis tests between the two supplement groups to answer the questions 
about which supplement type and dose effects growth the most. 

#### Null Hypothesis No. 5
There is no difference in the Mean growth rates between 0.5-1 (mg/day) dosage 
sub-group of the Ascorbic Acid and Orange Juice supplement groups. 


```{r chunk20}
#check rates and hypothesis test for groups according to dose & supplement
t.test(tg11$rate,tg21$rate,var.equal=F,paired=F)
```

It is clear from the large negative t-value that the means are not the same. 
The confidence interval does not contain 0 and is entirely negative implying 
that, according to this test, the average value of tooth growth in the first 
group (ascorbic acid) will lie below the average value of the tooth growth in 
the second group (orange juice). 

<b> <u> We can, therefore reject the null hypothesis no. 5</u> </b>

#### Null Hypothesis No. 6

There is no difference in the Mean growth rates between 2(mg/day) dosage 
sub-group of the Ascorbic Acid and Orange Juice supplement groups. 

```{r chunk21}
#check rates and hypothesis test for groups according to dose & supplement
t.test(tg12$rate,tg22$rate,var.equal=F,paired=F)
```

This result is more interesting. As can be seen from the plot, the average tooth
growth rate for
each supplement type begins to get close to one another once the dosage begins 
to reach the 2(mg/day) level. 
The t-value is small and p-value large, The confidence interval of the 
difference is small and does contain 0. 
We can conclude that based on evidence provided by this data there is 
no difference in the mean growth rates of the 2(mg/day) subgroup of each 
supplement group. 
<b> <u> We cannot, therefore, reject null hyposthesis no. 6. </u> </b>

It would be interesting to make one final test using all of the 30 
observations of each supplement group. 

We propose:  

#### Null hyposthesis no. 7.  
There is no difference in the Mean growth rates between the Ascorbic Acid and 
Orange Juice supplement groups. 

```{r chunk22}
#check rates and hypothesis test for groups according to supplement
t.test(tg1$rate,tg2$rate,var.equal=F,paired=F)
```

This result shows that overall if we take normalised growth rates for all 
observations in the two groups, the growth rates for each supplement type
is different as the t-value is large, p-value is negligible and the confidence
interval of the difference does not contain 0. It is safe to conclude that the 
mean growth rates for each supplement type is different. 

<b> <u> We can, therefore reject the null hypothesis no. 7</u> </b>


### Conclusions

After analysing the data and proposing seven hypothesis we can conclude that:  

1) Average growth rates of teeth for the animals that were given 0.5 to 1(mg/day) 
of ascorbic acid as a supplement are similar. The average growth rate of the 
teeth of animals given 2(mg/day) ascorbic acid as a supplement is not similar to 
the average growth rates of the group of animals given 0.5-1(mg/day) ascorbic 
acid as a supplement.  
2) Average growth rates of teeth for the animals that were given 0.5 to 1(mg/day) 
of orange juice as a supplement are similar. The average growth rate of the 
teeth of animals given 2(mg/day) orange juice as a supplement is not similar to 
the average growth rates of the group of animals given 0.5-1(mg/day) orange 
juice as a supplement.  
3) The average growth rate of teeth of animals given 0.5-1(mg/day) ascorbic acid 
supplement is not similar to the average growth rate of teeth of animals given 
0.5-1(mg/day) orange juice supplement.   
4) The average growth rate of teeth of animals given 2(mg/day) ascorbic acid 
supplement is similar to the average growth rate of teeth of animals given 
0.5-1(mg/day) orange juice supplement.  
5) Overall, if all data points are taken together (thereby ignoring 
conlcusion 4 above), 
the average growth rate of 
teeth of animals given ascorbic acid 
supplement is not similar to the average growth rate of teeth of animals given 
orange juice supplement.

