---
title: 'Project: Using the t distribution in hypothesis testing'
author: "Vineet W. Singh"
date: "19 February 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
The t distribution is used to test statistics on sets of data when the 
sample sizes are small. This brief project demonstrates the use of the t 
distriution in finding 
confidence intervals and conducting hypothesis tests on the ToothGrowth 
(included with R) dataset and this has 60 observations.  

# Analysis  
The ToothGrowth data set contains data regarding Vitamin C given as a
supplement (in two forms) and the resulting growth of teeth in 60 guinea pigs.  

This analysis will answer the following questions:  
1) Do the teeth grow the same, for all dosage of supplements or do different 
doses of supplements result in different growth.  
2) Do both the supplements result in the same rate of growth of teeth or is 
the growth rate different for each supplement.  
```{r chunk10, include=FALSE }
library(datasets)
library(dplyr)
```
The structure of the ToothGrowth dataset can be found using the str function.  
```{r chunk11, echo=FALSE}
str(ToothGrowth)
```
We need to partition the data according to the supplements concerned, either
"VC" for Ascorbic Acid or "OJ" for orange juice. Normalising length by dose and  
storing this as a new variable "rate" will help in the tests.  
```{r chunk12, echo=FALSE}
tg1 <- ToothGrowth %>% filter(supp=="VC")
tg1<-tg1 %>% mutate(rate=len/dose)
tg2 <- ToothGrowth %>% filter(supp=="OJ")
tg2<-tg2 %>% mutate(rate=len/dose)
```
**Summary Statistics for each group are:**  
__Ascorbic Acid : Summary for 10 observations in each subgroup__  
For 0.5(mg/day) subgroup - Mean: 16, SD: 5.49;  
For 1(mg/day) subgroup - Mean: 16.8, SD: 2.52;  
For 2(mg/day) subgroup- Mean: 13.1, SD: 2.40.  
__Orange Juice : Summary for 10 observations in each subgroup__  
For 0.5(mg/day) subgroup - Mean: 26.5, SD: 8.92;  
For 1(mg/day) subgroup - Mean: 22.7, SD: 3.91;  
For 2(mg/day) subgroup - Mean:13.0, SD: 1.33.  

An interesting plot (Appendix) is included showing the growth rate pattern for 
each supplement type. 

From the summary & plot it can be observed that for each subgroup,
the growth rates vary according to dosages given, with the maximum variabilty in 
the 0.5 mg group for both supplements and different growth rates for each 
supplement. 

## Assumptions
We will need to make some assumptions about the data before we carry out the 
hypothesis tests.  
We can assume that the growth rate in each animal in each subgroup is 
independent. Further more we can assume that the variance 
within each of the 6 subgroups (supplement & dosage) is the same.
However, it is probably not very wise to assume that the variance accross 
different subgroups is the same as, clearly the dosage levels are different 
across each subgroup in one supplement group.  

## Testing and Results   
**Null Hypothesis No. 1**  
Mean growth rates for the 0.5(mg/day) group and the 1(mg/day) group of the 
ascorbic acid set are the same.  

Since we only have 10 measurements each subgroup, we need to make 
confidence intervals and do tests using the t distribution. 
Results of the t-test (Appendix: Test 1) to check the hypothesis are 
t-value: -0.42396, p-value > .50 and the confidence interval (CI) for 
the difference of means contains 0. **We cannot, therefore, reject null**
**hypothesis no. 1. **  

**Null Hypothesis No. 2.**  
Mean growth rates for the 1(mg/day) group and the 2(mg/day) group of the 
ascorbic acid set are the same.  

Result of the t.test for this hypothesis (Appendix Test 2) shows a high t-value 
of 3.3662, a low p value and a 
confidence interval where the difference is well above 1. **We can, therefore,**
**reject the null hypothesis no. 2** and conclude that the means for 
the two subgroups are different.

**Null hypothesis No. 3:  **  
Mean growth rates for the 0.5(mg/day) group and the 1(mg/day) group of the 
orange juice set are the same.  
**Null hypothesis No. 4**  
Mean growth rates for the 1(mg/day) group and the 2(mg/day) group of the 
orange juice set are the same.  

We find from the results of test 3 (Appendix: Test 3/4) that the t-value is 
above 
1 but not very high, the CI of the difference contains 0 so we cannot 
rule out that the means might not be the same. Results of test number 4 tell a 
different story, however, the t-value is very large, the p-value is negligible 
and the confidence interval of the difference does not contain 0.  
**Using this argument, we cannot reject null hypothesis no. 3,**
**but we can reject null hypothesis no. 4.**  

Based on the results of the 4 hypothesis tests above, groups of 0.5(mg/day) and 
1(mg/day) each supplement types are merged and further tests are carried out 
for to test for growth between the two supplement types. 
There will be 3 further tests, one between the merged 0.5-1(mg/day) subgroup of 
each supplement type, one between the 2 (mg/day) subgroup of 
each supplement type and one test that compares all the data taken together for 
each supplement type.  

**Null Hypothesis No. 5**  
There is no difference in the Mean growth rates between 0.5-1 (mg/day) dosage 
sub-group of the Ascorbic Acid and Orange Juice supplement groups. 

Refering to the results of test no. 5 (Appendix: Test 5), we have  large 
negative t-value and this implies that the means for the two supplement types 
are not the same. The CI does not contain 0 and is entirely negative implying 
that, according to this test, the average value of tooth growth in the first 
group (ascorbic acid) will lie below the average value of the tooth growth in 
the second group (orange juice). **We can, therefore reject null**
**hypothesis no. 5**  

**Null Hypothesis No. 6**  
There is no difference in the Mean growth rates between 2(mg/day) dosage 
sub-group of the Ascorbic Acid and Orange Juice supplement groups. 

The result for test no. 6 (Appendix: Test 6) is interesting. 
This can also be inferred from the plot, that the average tooth
growth rate for each supplement type begins to get close to one another once 
the dose starts getting close to the 2(mg/day) level. 
The t-value is small, p-value large, The CI of the 
difference of means is small and does contain 0. 
We can conclude that based on evidence provided by this data there is 
no difference in the mean growth rates of the 2(mg/day) subgroup of each 
supplement group. **We cannot, therefore, reject null hyposthesis no. 6.**  

It would be interesting to make one final test using all of the 30 
observations of each supplement group. 

**Null hyposthesis no. 7.**  
There is no difference in the Mean growth rates between the Ascorbic Acid and 
Orange Juice supplement groups. 

The result of test 7 shows that overall if we take normalised growth rates for 
all observations in the two groups, the growth rates for each supplement type
is different as the t-value is large, p-value is negligible and the CI of the 
difference of the means does not contain 0. It is safe to conclude that the 
mean growth rates for each supplement type is different.  
**We can, therefore reject the null hypothesis no. 7.**  

# Conclusion  

After analysing the data and proposing seven hypothesis we can conclude that:  
1) Average growth rates of teeth for the animals that were given 0.5 to 1(mg/day) 
of ascorbic acid as a supplement are similar. The average growth rate of the 
teeth of animals given 2(mg/day) ascorbic acid as a supplement is not similar to 
the average growth rates of the group of animals given 0.5-1(mg/day) ascorbic 
acid as a supplement.  
2) Average growth rates of teeth for the animals that were given 0.5 to 1(mg/day) 
of orange juice as a supplement are similar. The average growth rate of the 
teeth of animals given 2(mg/day) orange juice as a supplement is not similar to 
the average growth rates of the group of animals given 0.5-1(mg/day) orange 
juice as a supplement.  
3) The average growth rate of teeth of animals given 0.5-1(mg/day) ascorbic acid 
supplement is not similar to the average growth rate of teeth of animals given 
0.5-1(mg/day) orange juice supplement.   
4) The average growth rate of teeth of animals given 2(mg/day) ascorbic acid 
supplement is similar to the average growth rate of teeth of animals given 
0.5-1(mg/day) orange juice supplement.  
5) Overall, if all data points are taken together (thereby ignoring 
conlcusion 4 above), 
the average growth rate of 
teeth of animals given ascorbic acid 
supplement is not similar to the average growth rate of teeth of animals given 
orange juice supplement.

# Appendix
**Ascorbic Acid : Summary  **  
```{r chunk13}
#calculate summaries for sub group 
tg1 %>% group_by(dose) %>% summarise(mean=mean(rate),sd=sd(rate),n=n())
```
**Orange Juice : Summary**  
```{r chunk14}
#calculate summaries for sub group 
tg2 %>% group_by(dose) %>% summarise(mean=mean(rate),sd=sd(rate),n=n())
```
**PLOT**
```{r chunk15}
plot(x=row.names(tg1),y=tg1$rate,type="p",pch=1,ylab="Rate",
     main="Tooth Growth Rate According to Supplement",
     xlab="Observation No.", ylim=c(0,50),col=1)
lines(x=row.names(tg1),y=tg1$rate,type="l",lty=1,lwd=2,col=1)
points(x=row.names(tg2),y=tg2$rate,pch=1,col=2)
lines(x=row.names(tg2),y=tg2$rate,type="l",lty=1,lwd=2,col=2)
legend("topright", c("Ascorbic Acid","Orange Juice"),col=c(1,2),lwd=4)
```
```{r chunk16}
tg11<- tg1 %>% filter(dose==0.5)
tg12<- tg1 %>% filter(dose==1)
tg13<- tg1 %>% filter(dose==2)

tg21<- tg2 %>% filter(dose==0.5)
tg22<- tg2 %>% filter(dose==1)
tg23<- tg2 %>% filter(dose==2)
```
**RESULTS OF TEST 1**
```{r chunk17}
t.test(tg11$rate,tg12$rate,paired=F,var.equal = F)
```
**RESULTS OF TEST 2**
```{r chunk18}
t.test(tg12$rate,tg13$rate,paired=F,var.equal = F)
```
**RESULTS OF TEST 3**  
```{r chunk19}
t.test(tg21$rate,tg22$rate,paired=F,var.equal = F)
```
**RESULTS OF TEST 4**  
```{r chunk20}
t.test(tg22$rate,tg23$rate,paired=F,var.equal = F)
```
```{r chunk 201, echo=FALSE}
#CI for mean growth rate for tg11 and tg12 includes 0 so we cannot rule out
#that the mean growth rate for both tg21 and tg22 can be the same
#it might make sense to group tg11 and tg12 together and assume a common mean 
#growth rate. 

#merge tg21 and tg22 into one frame
tg11<-rbind(tg11,tg12)

#reassign tg23 to tg22
tg12<-tg13

#CI for mean growth rate for tg21 and tg22 includes 0 so we cannot rule out
#that the mean growth rate for both tg21 and tg22 can be the same
#it might make sense to group tg21 and tg22 together and assume a common growth 
#rate. 

#merge tg21 and tg22 into one frame
tg21<-rbind(tg21,tg22)

#reassign tg23 to tg22
tg22<-tg23
```
**RESULTS OF TEST 5**
```{r chunk21}
#check rates and hypothesis test for groups according to dose & supplement
t.test(tg11$rate,tg21$rate,var.equal=F,paired=F)
```
**RESULTS OF TEST 6**
```{r chunk22}
#check rates and hypothesis test for groups according to dose & supplement
t.test(tg12$rate,tg22$rate,var.equal=F,paired=F)
```
**RESULTS OF TEST 7**
```{r chunk23}
#check rates and hypothesis test for groups according to supplement
t.test(tg1$rate,tg2$rate,var.equal=F,paired=F)
```
